---
title: "Data Analysis"
output: 
  html_document: 
    keep_md: yes
    toc: true
    toc_float: true
---

## About

Statistical analysis performed as a part of research published in a medical research journal. 

* N.V. Halinouskaya, _S.V. Samsonau_, N.N. Usava, V.B. Smychek, “Thyroid Hormone Status in Stroke and Transient Ischemic Attack Patients”, Neurologijos Seminarai, Volume 19, Issue 3, 207, 2015

* The paper can be found on the [website of the journal](http://www.neuroseminarai.lt/wp-content/uploads/2017/02/Neuro_2015_Nr3_207-209.pdf)

* [Return](https://github.com/ssamsonau/Thyroid_hormone_status) to GitHub project page



## Load data

```{r}
library(data.table)
library(ggplot2)

#open file
DT <- fread("./dataPreparation//data_repaired_translitarated.csv", colClasses=rep("character", 977))
#dim(DT)

#choose coloumns needed
chosen.col <- c(184, 185, 186, 28, 16, 6) 
print("Names of chosen coloumns are:")
names(DT)[chosen.col]

#change names to readable
DTsub <- subset(DT, select=chosen.col) 
setnames(DTsub, c("T4", "TSH", "ATPO", "group", "age", "gender"))
print("changed to :")
names(DTsub)

#variables for plots
variable.names <- c("T4", "TSH", "ATPO")
var.plot.notation <- c("T4, pm/l", "TSH, mIU/L", "Anti-TPO, IU/mL")
```

## Prepare data

```{r, results='hide'}
#make data numeric
num.var <- c("T4", "TSH", "ATPO", "age")
DTsub[, eval(num.var):=lapply(.SD, as.numeric), .SDcols=num.var]

#make appropriate factor levels
group.values <- c("1", "2", "3", "90", "7")
new.group.values <- c("LS", "TS", "TIA", "control", "control")

library(plyr)
DTsub[, group:=mapvalues(group, from=group.values, to=new.group.values)]
DTsub <- DTsub[group %in% new.group.values, ]
DTsub[ , group := factor(group)]
DTsub[ , group := factor(group, levels(group)[c(1, 4, 2, 3)])]

#gender
DTsub[, gender:=mapvalues(gender, from=c("m", "zh"), to=c("male", "female"))]

#age
DTsub[age<=45, age.cohort:="young"]
DTsub[age>45, age.cohort:="elderly"]
DTsub <- DTsub[age.cohort %in% c("young", "elderly"), ]

#all data for young, for transient ischemic attack are NA except one that is 0. So we do not use them
DTsub <- DTsub[! (group=="TIA" & age.cohort=="young"), ]

#use only the data where all records available and refactor
DTsub <- na.omit(DTsub)
DTsub[, gender:=factor(gender)]
DTsub[, age.cohort:=factor(age.cohort)]

```

```{r}
print.overview <- function(dt){#function to print overview of data
    #summary(DTsub)
    print("data summary---")
    print( summary(dt) )
        
    print("Male and Female separatelly---")
    data.overview <- dt[, list(.N, mean(age), sd(age)), by=list(age.cohort, group, gender)]
    setnames(data.overview, c("group", "V2", "V3"), c("group", "age.mean", "age.sd"))
    setkey(data.overview, age.cohort, group, gender)
    print( unique( data.overview ) )

    print("Male and Female together---")
    data.overview <- dt[, list(.N, mean(age), sd(age)), by=list(age.cohort, group)]
    setnames(data.overview, c("group", "V2", "V3"), c("group", "age.mean", "age.sd"))
    setkey(data.overview, age.cohort, group)
    print( unique( data.overview ) )

    }

print.overview(DTsub)
```


## Plot

### With outliers
```{r plots before removing outliers}
give.n <- function(x){ #function for positioning numbers on plots
  return(data.frame(y = max(x), label = paste0("       ",length(x))))
}

for(variable in variable.names){
    plot.var <- ggplot(data=DTsub,
                       aes(x=group, y =get(variable), fill=group)) + geom_boxplot() +
        facet_grid(.~ age.cohort, drop=T, space="free_x") +
        theme_bw(16) +
        theme(legend.position="none", axis.title.x = element_blank()) + 
        ylab(var.plot.notation[variable==variable.names]) +
        stat_summary(fun.data = give.n, geom = "text") +
        scale_fill_grey(start = 0.5, end = .9)

    print(plot.var)

    tiff(paste0("outfile",variable,"wo.out.tiff"), res=300, height=5, width=6.80, units="in")
    print(plot.var)
    dev.off()    
}

```
 
 
### Without outliers
Here we are cuttiong out records for which values of variables are 5 IQR (Interquartile ranges) lower or higher relative to 1st and 3rd quartiles correspondingly.

```{r, results='hide'}
remove_outliers <- function(x, na.rm = TRUE, countIQR = 5, ...) {
    H <- countIQR * IQR(x, na.rm = na.rm)
    qrts <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
    y <- x
    y[x < (qrts[1] - H)] <- NA
    y[x > (qrts[2] + H)] <- NA
    y
}

DTsub.wout <- data.table(DTsub)

DTsub.wout[age.cohort=="young", eval(variable.names):=lapply(.SD, remove_outliers), .SDcols=variable.names]
DTsub.wout[age.cohort=="elderly", eval(variable.names):=lapply(.SD, remove_outliers), .SDcols=variable.names]

DTsub.wout <- na.omit(DTsub.wout)
```

```{r}
print.overview(DTsub.wout)
```

```{r plots without outliers}
theme_set(theme_grey(base_size = 18))
for(variable in variable.names){
    
    plot.var <- ggplot(data=DTsub.wout,
                       aes(x=group, y =get(variable), fill=group)) + geom_boxplot() +
        theme_bw(16) +
        facet_grid(.~ age.cohort, drop=T, space="free_x") +
        theme(legend.position="none", axis.title.x = element_blank()) + 
        ylab(var.plot.notation[variable==variable.names]) +
        stat_summary(fun.data = give.n, geom = "text")+
        scale_fill_grey(start = 0.5, end = .9)
    
    print(plot.var)

    tiff(paste0("outfile",variable,"wo.out.tiff"), res=300, height=5, width=6.80, units="in")
    print(plot.var)
    dev.off()
}
    
```


## Statistics (outliers removed)

First let us see if all groups are identical. In order to do this we could use ANOVA test if  

 * the data would be independent (yes in this case)
 * normally distributed (no in this case). ANOVA can be robust to not normal data if sample sizes are equal (no in this case).
 * have a identical variances (no for elderly patients - checked bellow with Levene's Test).

```{r}
print.test <- function(dt, age.cohort.v, variable, statistics.v, test="kruskal"){
        print(age.cohort.v)
        print(variable)
        DTtemp <- na.omit(dt[age.cohort==age.cohort.v, list(get(variable), group)])
        setnames(DTtemp, "V1", eval(variable))
        
        DTstat <- DTtemp[, list(.N, mean(get(variable)), var(get(variable)), median(get(variable))), 
                         by=group]
        setnames(DTstat, c("N", "V2", "V3", "V4"), c("# of observations", "mean", "variance", "median"))
        
        if(test == "levene"){
            library(car)
            print( leveneTest(get(variable) ~ group, data=DTtemp) )
        }
        if(test == "kruskal"){
            print( DTstat  )    
            print(  kruskal.test(get(variable) ~ group, data=DTtemp)  )
        }
}
for(age.cohort.v in c("elderly", "young")){
    for(variable in variable.names){
        print.test(DTsub.wout, age.cohort.v, variable, test="levene")        
    }    
}
```

Thus, in the case we should use Kruskal-Wallis test, which does not require from data to satisfy those conditions.

```{r}
for(age.cohort.v in c("elderly", "young")){
    for(variable in variable.names){
        print.test(DTsub.wout, age.cohort.v, variable, test="kruskal")        
    }    
}


```

We see there is at least one coulomn is different (p value < 0.05) for

 * elderly and T4
 * young and ATPO

Now we can use paired comparassion to find what exactly is statistically different inside these groups.
In order to do this we use post-hoc comparassion following Siegel and Castellan procedure.

### Elderly cohort
```{r}
library(pgirmess)
kruskalmc(T4 ~ group, data=DTsub.wout[age.cohort=="elderly", ])
#kruskalmc(T4 ~ group, data=DTsub.wout[age.cohort=="elderly", ], cont='two-tailed')

```
We see strong evidence of a difference between

 * TIA and control
  
### Young cohort
```{r}
kruskalmc(ATPO ~ factor(group), data=DTsub.wout[age.cohort=="young", ])
#kruskalmc(ATPO ~ factor(group), data=DTsub.wout[age.cohort=="young", ], cont='two-tailed')

```

We see here strong evidence of a difference between

 * TS and control
 * LS and control